Versioned data: why it’s needed and how it can be achieved (easily and cheaply)

### 2017-08-03 ms.tex
some text cleanup

Data is now a first-class scientific product, as is evidenced by the advent of this journal \cite{Editorial-2014}. Increasingly, funding bodies, publishers, and scientific social norms are both requiring data to be published alongside research publications and recognizing the value of producing and sharing data without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}. This has been a rapid, and exciting, sea change and the scientific community is still grappling with how best to disseminate data and credit data creators and aggregators \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}. One thing that is clear is that there is no single best solution for managing and distributing data, as data types and use is incredibly heterogeneous. At one end of the spectrum are massive, centralized repositories that are constantly being added to; this is a great solution to storing and distributing relatively standardized data such as sequences (e.g., \texttt{GenBank}, \texttt{SRA}) (Table \ref{tab:sql_v_versioneddata}). At the other end are repositories for storing a wide variety of static, stand-alone datasets (eg. \texttt{Dryad}, \texttt{Figshare}, and \texttt{Zenodo}; or less optimally, as supplements to research papers)(Table \ref{tab:sql_v_versioneddata}). We suggest there is a substantial middle ground that is not well served by either of these existing paradigms, where a lightweight system for serving versioned datasets is needed.

In this article we i) introduce the concept of Versioned Data Delivery (\textsc{vdd}) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new \textsc{R} package called \texttt{datastorr}, an implementation of a \textsc{vdd} system. While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms. While the examples cited here are in ecology and evolution; the issues identified and the solutions put forward are relevant to nearly all data-oriented scientific domains.

@@

Almost all useful databases are living--they \emph{should} get better
through time either by the addition of new data or by improving the
quality of existing data. For example a dataset on plant characteristics
might become outdated as taxonomic changes occur. Updating the dataset might only
require a small amount of work, but if there is no
way to distribute the updated dataset, then that work has to be re-done by
every user of that dataset. Moreover, those required changes add steps
between the canonical (or master) dataset and the new analysis (and publication),
making reproducibility of the new analysis more difficult than it needs
to be. If the canonical dataset is versioned--and versions are
accessible--the up-to-date data can be distributed, used, and analyses
based on any of the released versions data can be reproduced.

@@

Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development.  Essentially version control tracks line-by-line changes in text files and creates and maintains a history of those changes. Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016}. Git is attractive for data management because it tracks all changes in monitored files. This history is visible to anyone interacting with the repository. It also allows users to annotate changes (`commits') with informative messages detailing the rationale for those changes.

@@

\subsection{Conclusion}

>>>>

Data is now a first-class scientific product, as is evidenced by the advent of this journal (\cite{Editorial-2014}). Increasingly, funding bodies, publishers, and scientific social norms are both requiring data to be published alongside research publications and recognizing the value of producing and sharing data without any accompanying analyses \cite[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}. This has been a rapid, and exciting, sea change and the scientific community is still grappling with how best to disseminate data and credit data creators and aggregators \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}. One thing that is clear, however, is that there is no single best solution for managing and distributing data, as data types and use is incredibly heterogeneous. At one end of the spectrum are massive, centralized repositories that are constantly being added to; this is a great solution to storing and distributing relatively standardized data such as sequences (e.g., \texttt{GenBank}, \texttt{SRA}) (Table \ref{tab:sql_v_versioneddata}). At the other end are repositories for storing a wide variety of static, stand-alone datasets (eg. \texttt{Dryad}, \texttt{Figshare}, and \texttt{Zenodo}; or less optimally, as supplements to research papers) (Table \ref{tab:sql_v_versioneddata}). We suggest there is a substantial middle ground that is not well served by either of these existing paradigms, where a lightweight system for serving versioned datasets is needed.

In this article we i) introduce the concept of Versioned Data Delivery (\textsc{vdd}) and describe why it is needed; ii) outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and iii) introduce a new \textsc{R} package called \texttt{datastorr}, an implementation of a \textsc{vdd} system. While we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms. Likewise, the examples cited here come from our own research areas, ecology and evolutionary biology, but we argue that the issues we have identified and the solutions we have put forward are relevant to nearly all data-oriented scientific domains.

@@

Almost all useful databases are living--they \emph{should} get better
through time either by the addition of new data or by improving the
quality of existing data. For example a dataset on plant characteristics
might become outdated as taxonomic changes occur. Updating the dataset might only
require a small amount of work, but if there is no
way to distribute the updated dataset, then that work has to be re-done by
every user of that dataset. Moreover, those required changes add steps
between the canonical (or master) dataset and the new analysis (and publication),
making reproducibility of the new analysis more difficult than it needs
to be. If the canonical dataset is versioned--and versions are
accessible--the up-to-date data can be distributed, used, and analyses
based on any of the previously released versions data can be reproduced.

@@

Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development.  Essentially version control tracks line-by-line changes in text files and creates and maintains a history of those changes. Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016}. \texttt{git} is attractive for data management because it tracks all changes in monitored files. This history is visible to anyone interacting with the repository. It also allows users to annotate changes (`commits') with informative messages detailing the rationale for those changes.

@@

%\subsection{Conclusion}
